---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r}
library(dplyr)

library(gapminder)
data("gapminder")
attach(gapminder)

head(gapminder)


```
I've loaded the gapminder data and now I am going to manipulate it with dplyer.

```{r}
mean(gapminder$gdpPercap[continent == "Africa"])

mean.gdpAfrica <- mean(gapminder$gdpPercap[continent == "Africa"])
mean.gdpAfrica
mean.gdpAsia <- mean(gapminder$gdpPercap[continent == "Asia"])
mean.gdpAsia
mean.gdpAmericas <- mean(gapminder$gdpPercap[continent == "Americas"])
mean.gdpAmericas


```

### Working with data the dplyr

select() - subset variables (columns)
filter() - subset observations (rows)
group_by() - counts and groups data
summarize() - useful information about the data
mutate() - add a new column based on the data

```{r}

ncol(gapminder)
names(gapminder)
subset_gap <- gapminder %>% select(year, country, gdpPercap)
subset_gap
dim(subset_gap)


```


filter()

```{r}
year_country_gdp_euro <- gapminder %>%
  filter(continent == "Europe") %>%
  select(year, country, gdpPercap)

year_country_gdp_euro


```

Socrative questions

```{r}
africa_2007 <- gapminder %>%
  filter(continent == "Africa" & year == 2007) %>%
  select(year, country, lifeExp)

dim(africa_2007)
```


group_by() function subset your data by chunking it based on a column

```{r}

str(gapminder)

gap_by_continent <- gapminder %>% group_by(continent)

head(gap_by_continent)
str(gap_by_continent)




```

group_by() turns a single data frame into a set of data frames grouped based on some variable.


```{r}
gdp_by_continent <- gap_by_continent %>%
  summarise(mean_gdpPercap = mean(gdpPercap))

gdp_by_continent

```


Compute average life expectancy across all African countries by year. In how many years did average African life expectancy decrease?

```{r}

life_africa <- gapminder %>%
  filter(continent == "Africa") %>%
  group_by(year) %>%
  summarise(mean_life = mean(lifeExp))

life_africa


```

Calculate the average life expectance er country. Which has the longest average life expectancy? Which has the shortest average life expectancy?

```{r}
life <- gapminder %>%
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp))
life

life %>% filter(mean_lifeExp == min(mean_lifeExp))

life %>% filter(mean_lifeExp == max(mean_lifeExp))


```
mutate() is a function in dplyr that lets you add a column to a data frame based on the data.

```{r}
billgdp <- gapminder %>%
  filter(year == 2007) %>%
  mutate(billion_gdp = gdpPercap * pop / 10^9) %>%
  select(continent, country, billion_gdp)

colnames(billgdp)
dim(billgdp)


```

```{r}

gdp_by_cont <- gapminder %>%
  group_by(continent, year) %>%
  summarise(mean_gdp = mean(gdpPercap), 
            sd_gdp = sd(gdpPercap), 
            mean_pop = mean(pop), 
            sample_size = n(),
            se_gdp = sd_gdp / sqrt(sample_size))
gdp_by_cont

gdp_by_cont <- gdp_by_cont %>% data.frame()

str(gdp_by_cont)


```

### Combining dplyr with ggplot2

```{r}
library(ggplot2)

gapminder %>% filter(continent == "Asia") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()


```

```{r}

gap_wide <- read.csv("data/gapminder_wide.csv", stringsAsFactors = FALSE)


```

This data is in a wide format. We want to change it to a long format.

```{r}
head(gap_wide)

## To make the data long, we will use the function gather()

gap_long <- gap_wide %>%
  gather(obstype_year, obs_values, -continent, -country)
dim(gap_long)


```

Separate out measurement type from year

```{r}
gap_long<- gap_long %>% separate(obstype_year, into = c("obs_type", "year"), 
                                 sep = "_")

head(gap_long)

gap_long$year <- as.integer(gap_long$year)
head(gap_long)

```

Summarize the mean life expectancy by continent


```{r}
life_by_cont <- gap_long %>%
  filter(obs_type == "lifeExp") %>%
  group_by(continent) %>%
  summarize(mean_life = mean(obs_values))
life_by_cont

head(gap_long)

unique(gap_long$obs_type)


```


Turn long table into a more human-readable format

```{r}
gap_normal <- gap_long %>% spread(obs_type, obs_values)
str(gap_normal)
colnames(gap_normal)

gap_normal <- gap_normal[,names(gapminder)]

all.equal(gap_normal, gapminder)

gap_normal <- gap_normal %>% arrange(country, continent, year)

all.equal(gapminder, gap_normal)


```
